---
title: 'STAT656: Homework 1'
subtitle: Analyzing Airline Flight Delays.  
output:
  html_document: default
---

# Introduction and Data
We will be analyzing airline flight delays.  Note 
that this Rmd file will not knit in its current form.  It is necessary to complete some of the below items
for it to knit successfully.

## Download 

The first step in this assignment will be downloading the
flight data from the [Bureau of Transportation Statistics](https://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236&DB_Short_Name=On-Time)

Here are some instructions:

* Set "Filter Year" to 2019
* Set "Filter Period" to January
* Select the following fields:
  - Month
  - DayofMonth
  - DayofWeek
  - FlightDate
  - Reporting_Airline
  - DOT_ID_Reporting_Airline
  - OriginAirportID
  - OriginCityMarketID
  - Origin
  - OriginCityName
  - OriginState
  - DestAirportID
  - DestCityMarketID
  - Dest
  - DestCityName
  - DestState
  - DepDelay
  - DepDel15
  - ArrDelay
  - ArrDel15
  - Cancelled
  - CancellationCode

* Click "Download"
* Change the "Filter Period" to February and Click "Download"
* Change the "Filter Period" to March and Click "Download"
* ...
* Change the "Filter Period" to December and Click "Download"

## Uncompress and rename

Now, go to wherever your operating system defaults to downloading, more than likely your 'download' directory.
There will be 12 files .zip files there.  Uncompress each of these files and rename the resulting .csv file a new name corresponding
to the order in which they were downloaded (and hence the month of the year).  Using 'snake-case' would probably work well here

* 2019_01.csv
* 2019_02.csv
* ...
* 2019_12.csv

## Move the files

Make a subdirectory of your R working directory (the working directory is where your .Rmd file is located) called "database".  Now,
we want to run the following code to generate an R object that is a single flat file containing all the flights for 2019.

## Let's load in some packages.  

```{r loadPackages}
if(!require(tidyverse)){install.packages('tidyverse');require(tidyverse)}
```


# Problem 0 (20 pts)

Now, we want to take the data and make a data set. First, let's get the file names from the database directory

```{r getFileNames, cache = TRUE, cache.extra = file.info('./databases')}
paths = dir("database",pattern="2019_",full.names=TRUE)
print(paths)
names(paths) = basename(paths)
```

#### Answer 0.1
What does 'basename' do here?

'basename' removes the path up to the last path separator and only outputs the filename.If we take, "database/flightdelays2019_01.csv", it will output "flightDelays2019_01.csv". 

## Creating the data set

This next step can take a few minutes.  Let's time it! Accurately timing code is 
extremely delicate and controversial because it depends on all the settings on your computer, what
other programs you have running, architecture (do you have multiple cores/processors/...) 
even how long it has been since you rebooted!  

But, a simple and 
useful way is via `proc.time', which records the computer time of your processor.  Let's read in
each .csv file into an R data file that has all of 2019 flights. 

(note: we will use 'cache = TRUE'
so that R studio won't re-compile this section every time you knit.  It will detect any changes
to this `chunk' and re-run if there are any.  However, it doesn't detect if there are changes in
another chunk that affect this chunk.  For that, we need 'dependson'. In this case, a change
to the directory database would affect 'df'.  So, we want to add 'dependson' pointed to that chunk.Note that
each dependson chunk must be cached as well)

```{r createDataSet, cache = TRUE, dependson = 'getFileNames'}
srt    = proc.time()[3]
dfList = lapply(paths,read.csv)
df     = do.call(rbind,dfList)
end    = proc.time()[3]
cat('total time: ',end-srt)
df$X = NULL #get rid of that empty column
save(df, file="2019flights.Rdata")

totalTime = end - srt #### Answer 0.2 ## Overwrite the totalTime object so that it has the correct elapsed time
```

Answer: This process took `r totalTime` seconds.

## The data set

Let's look at some attributes of this data set

```{r dataSetSize}
dataSetSize = format(object.size(df),units='MiB')
```

#### Answer 0.3

The total size of the flights data set for 2019 is `r dataSetSize`

# Problem 1 (20 pts)

Some of the flights were cancelled and hence never flew.  These are recorded via the 'cancelled' feature.
How many total flights are there in the data set? How many were cancelled?  

Additionally, there are some missing values, coded as NA, in the data set.  Generally, we would want to
impute these missing values.  However, we will return to imputation later in the
semester.  For now, let's remove the rows (flights) with missing values. Make a new object called flightsNotCancelled that
have the flights that were not cancelled and have no missing values.  Make sure to use this for the rest of the assignment.  Lastly,
how many flights do we have remaining in flightsNotCancelled?

```{r problem1}
totalNum  = nrow(df)
cancelNum =  df %>%
  group_by(CANCELLED) %>%  #### Answer 1.1 ## Insert the correct group_by statement here
  summarise(n = n())

#I just want the integer, not the tibble:
cancelNum = as.integer(cancelNum[2,2])

flightsNotCancelled = df %>%
  filter(!CANCELLED == 1) %>% #### Answer 1.2  Insert the correct filter statement here
  na.omit

flightsNotCancelledCount = nrow(flightsNotCancelled)
```

#### Answer 1.3

The total number of flights is `r totalNum`.  

The total number of cancelled flights is `r cancelNum`. 

The total number of flights remaining after removing NAs is `r flightsNotCancelledCount`.

# Problem 2 (15 pts)

How many flights were there for each day of the week?

```{r problem2}
flightsPerDay = flightsNotCancelled %>% 
  group_by(DAY_OF_WEEK) %>% 
  summarize(total = n())

flightsPerDay

ggplot(data = flightsPerDay) + geom_point(aes(x = DAY_OF_WEEK, y = total/sum(total)),color='blue')
#### Answer 2.1 ## Alter this plot so that the points are blue and the plot is of the proportion of flights each day instead of total
```

# Problem 3 (20 pts)

Which month has the greatest proportion of late flights?

```{r problem3}
props = flightsNotCancelled %>% 
  mutate(delayInd = ifelse(DEP_DELAY > 0,1,0)) %>% 
  group_by(MONTH) %>% #### Answer 3.1 Fill in the missing part
  summarize(total = n(),success = sum(delayInd),prop = mean(delayInd))

props

greatPropIndex   = which.max(props$prop)
lateFlightsMonth = props$MONTH[greatPropIndex]
```

#### Answer 3.2

The month with the greatest proportion of late flights is `r lateFlightsMonth`

# Problem 4 (25 pts)

Which departure and arrival airport combination is associated with the worst median departure delay?  

```{r problem4}
medianDepDelayCombination = flightsNotCancelled  %>%
  select(ORIGIN,DEST,DEP_DELAY) %>% group_by(ORIGIN,DEST) %>%
  summarise_each(list(median=median))
names(medianDepDelayCombination) = cbind("ORIGIN","DEST","medianDepDelay")
combination = medianDepDelayCombination[which.max(medianDepDelayCombination$medianDepDelay),]
combination
```

#### Answer 4.1
The worst departure/arrival airport combination in terms of median delay is `r combination$ORIGIN`/`r combination$DEST` 

